Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
further revise language on gains & losses (#48)
#48

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.

@@

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.

@@

```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss. Most, but not all, EFDB-relevant ForC variables are shown here. Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused). Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon. **This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

>>>>

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.

@@

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.

@@

```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux. Most, but not all, EFDB-relevant ForC variables are shown here. Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused). Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon. **This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
further revise language on gains & losses (#48)
#48

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned net emissions reductions
under the 2015 Paris Agreement. However, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accessible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
inventories if made available through the Emission Factor Database
(EFDB) of the International Panel on Climate Change (IPCC). Here, we
develop and document a process for semi-automated submission of data
from ForC into the EFDB, assess the data available and submitted to
date, and provide recommendations for improving forest data collection,
analysis, and reporting to improve inventories of forest-sector
greenhouse gas emissions and removals. We begin by reconciling
terminology and mapping ForC fields into EFDB. This process required
some updates to the ForC database structure, leading to the release of a
new version of ForC (v4.0; described here). As of June 09, 2023, ForC
contained \textasciitilde19316 independent records relevant to EFDB,
1438 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. In the future,
forest C estimates in EFDB can be improved through targeted research to
fill critical gaps, reporting of information required by IPCC, and
continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests,
timely reporting of recent estimates will be especially critical to
accurate forest C inventories.

@@

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); t C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (described in section 2.1), plus any harvested wood products. For
each pool, \(\Delta C\) may be calculated using the ``Gain-Loss
Method'', which takes the difference between gains and losses, or using
the ``Stock-Difference Method'', which computes \(\Delta C\) based on C
stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
variables relevant to the IPCC methodology and to EFDB include C stocks,
net annual increments, and fluxes in the IPCC-defined pools.

@@

\begin{figure}
\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss. Most, but not all, EFDB-relevant ForC variables are shown here. Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused). Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon. **This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.}\label{fig:fig_variable_mapping}
\end{figure}

>>>>

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned net emissions reductions
under the 2015 Paris Agreement. However, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accessible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
inventories if made available through the Emission Factor Database
(EFDB) of the International Panel on Climate Change (IPCC). Here, we
develop and document a process for semi-automated submission of data
from ForC into the EFDB, assess the data available and submitted to
date, and provide recommendations for improving forest data collection,
analysis, and reporting to improve inventories of forest-sector
greenhouse gas emissions and removals. We begin by reconciling
terminology and mapping ForC fields into EFDB. This process required
some updates to the ForC database structure, leading to the release of a
new version of ForC (v4.0; described here). As of June 09, 2023, ForC
contained \textasciitilde19316 independent records relevant to EFDB,
1438 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
or out of the IPCC-defined C pools. Geographic representation is also
quite uneven, with the highest densities of relevant records in
temperate forests, and with relatively scant representation of tropical
forests in Africa and Asia. In the future, forest C estimates in EFDB
can be improved through targeted research to fill critical gaps,
reporting of information required by IPCC, and continued submission of
data from scientific publications to the EFDB. Given that climate change
is rapidly impacting the world's forests, timely reporting of recent
estimates will be especially critical to accurate forest C inventories.

@@

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); t C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (described in section 2.1), plus any harvested wood products.
Thus, C cycle variables relevant to the IPCC methodology and to EFDB
include C stocks, net annual increments, and fluxes in the IPCC-defined
pools.

@@

\begin{figure}
\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux. Most, but not all, EFDB-relevant ForC variables are shown here. Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused). Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon. **This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.}\label{fig:fig_variable_mapping}
\end{figure}
